data_proprocessing

### 2021-02-02 code/data_load.Rmd

---
title: "data_proprocessing"
author: "Hannah Sophie Heinrichs"
date: "1/20/2021"
output: 
  pdf_document: default
  html_document: default
---

@@

```{r "Package loading", include=FALSE}
library(lme4)
library(ggplot2)
library(haven)
library(plyr)
library(dplyr)
library(tidyr)
library(psych)
library(MuMIn)
```

### Loading Demographics
We get BMI from `ADI_MRT_Messueberblick_2019_04.csv` (this is an older version with correct columns (compared to _AW.xlsx). See README in `/data/p_02161/ADI_studie/metadata/`.

@@

```{r "Load data", include=FALSE}
bmi = read.csv(
  "/data/p_02161/ADI_studie/metadata/ADI_MRT_Messueberblick_2019_04_10.csv",
  na.strings = c("", "NA"),
  stringsAsFactors = FALSE
)
bmi_tmp = bmi[, c("X.1",
                  "BMI",
                  "X.2",
                  "BMI.1",
                  "X.3",
                  "BMI.2",
                  "Geschlecht",
                  "Alter")]

# collect subject IDs in one column
bmi_tmp$subj.ID = bmi_tmp$X.1  
bmi_tmp$subj.ID[!is.na(bmi_tmp$X.2)] = bmi_tmp$X.2[!is.na(bmi_tmp$X.2)]  
bmi_tmp$subj.ID[!is.na(bmi_tmp$X.3)] = bmi_tmp$X.3[!is.na(bmi_tmp$X.3)]

bmi_tmp = bmi_tmp[, c("subj.ID", "BMI", "BMI.1", "BMI.2", "Geschlecht", "Alter")]

# get BMI data from another ADI table which contains more subjects
adi_data = read_sav("/data/p_02161/ADI_studie/metadata/Daten ADI aktuell.sav")
# select variables: only BMI
bmi_repl = adi_data[, c(
  "ADI_Code",
  "BMI_BL",
  "BMI_6m",
  "BMI_12m",
  "Sex",
  "Age_BL",
  "Age_6m",
  "Age_12m",
  "Educa_BL"
)]
bmi_repl$subj.ID = gsub("-", "", bmi_repl$ADI_Code)

# replace some missing BMI in the "Messueberblick" table with values from ADI_aktuell.sav
abmi = merge(bmi_tmp, bmi_repl, by = "subj.ID", all = TRUE)
abmi$BMI_BL[is.na(abmi$BMI_BL)] = abmi$BMI[is.na(abmi$BMI_BL)]
abmi$BMI_6m[is.na(abmi$BMI_6m)] = abmi$BMI.1[is.na(abmi$BMI_6m)]
abmi$BMI_12m[is.na(abmi$BMI_12m)] = abmi$BMI.2[is.na(abmi$BMI_12m)]
abmi$Sex[is.na(abmi$Sex)] = abmi$Geschlecht[is.na(abmi$Sex)]
abmi$Age_BL[is.na(abmi$Age_BL)] = abmi$Alter[is.na(abmi$Age_BL)]

# clean up data frame
lbmi <- abmi %>% gather(tp, BMI, BMI_BL, BMI_6m, BMI_12m)
lbmi = lbmi[, c("subj.ID", "Age_BL", "Sex", "tp",  "BMI")]
lbmi$tp = as.factor(lbmi$tp)
levels(lbmi$tp) = c("fu2", "fu", "bl")
lbmi$tp <- factor(lbmi$tp, levels = rev(levels(lbmi$tp)))
lbmi$Sex = as.factor(lbmi$Sex)
levels(lbmi$Sex) = c("male", "female", "male", "female")
```

```{r "Compare abmi with bmi from QOL list", include=FALSE }
# compare abmi (Messueberblick, ADI_aktuell.sav) with bmi from QOL list for those subjects who have it
QOL_sheet = read.csv(
  "/data/pt_02161/Results/Project1_headmotion/QOL_sheet_of_ADI_Messueberblick.csv"
)
QOL_sheet$subj.ID = gsub("-", "", QOL_sheet$X.1)

@@

bmi_compare = merge(QOL_sheet, abmi, by = "subj.ID", all = TRUE)
sum(bmi_compare$BMI.BL - bmi_compare$BMI_BL, na.rm = T)
sum(bmi_compare$BMI.FU - bmi_compare$BMI_6m, na.rm = T)
sum(bmi_compare$BMI.FU_2 - bmi_compare$BMI_12m, na.rm = T) # because of ADI061
```

@@

README: "Unfortunately, there are errors regarding BMI assignment (e.g. `/data/p_02161/ADI_studie/metadata/ADI_MRT_Messueberblick_AW.xlsx`, first sheet, where ADI002's FU2 weight is assigned to BL, for example). Similarly, group coding is wrong in the first two sheets (`ADI_MRT_Messueberblick_2019_04_` and `summary with SPSS data`). Correct group information, based on BMI trajectory, seems to be in sheet `QOL` (see ADI080, ADI036).
"

@@

```{r "Write group coding for all subjects", echo=FALSE}
bmi_group_recoded <- bmi_group[!(bmi_group$subj.ID == "ADI061" &
                         bmi_group$tp == "bl"),]
# recode fu and fu2 into bl and fu
bmi_group_recoded$tp[(bmi_group_recoded$subj.ID == "ADI061" &
                         bmi_group_recoded$tp == "fu")] = "bl"
bmi_group_recoded$tp[(bmi_group_recoded$subj.ID == "ADI061" &
                         bmi_group_recoded$tp == "fu2")] = "fu"

write.csv(bmi_group_recoded[bmi_group_recoded$tp=="bl",], "/data/p_02161/ADI_studie/metadata/group_list_all.csv")
# as all other tables have the original coding for tp in ADI061, we have wo continue working with bmi_group and make the same recoding in the very end again.
```

@@

These infos were gathered during data copying, organizing and preprocessing, especially of the rs data. Yet, infos from other sequences are also included. See `/data/pt_02161/Analysis/Preprocessing/qa/README.md` for further information.
```{r "Loading rs QA infos from /data/pt_02161/Analysis/Preprocessing/qa/rs_qa/ADI_rs_qa.csv", echo=FALSE}
rs_QA=read.csv("/data/pt_02161/Analysis/Preprocessing/qa/rs_qa/ADI_rs_qa.csv")
for (i in 1:nrow(rs_QA)){
  #print(strsplit(toString(rs_QA[i,1]),'_')[[1]][1])
  rs_QA[i,"subj.ID"]=strsplit(toString(rs_QA[i,1]),'_')[[1]][1]
  rs_QA[i,"tp"]=strsplit(toString(rs_QA[i,1]),'_')[[1]][2]
}

@@

See `/data/pt_02161/Analysis/Preprocessing/qa/README.md` for further information.
```{r "Loading FS infos from /data/p_02161/ADI_studie/BIDS/derivatives/freesurfer/QA/freesurfer_qa_qoala_11.7.csv", echo=FALSE}
FS_QA=read.csv("/data/p_02161/ADI_studie/BIDS/derivatives/freesurfer/QA/freesurfer_qa_qoala_11.7.csv")
for (i in 1:nrow(FS_QA)){
  #print(strsplit(toString(rs_QA[i,1]),'_')[[1]][1])
  FS_QA[i,"subj.ID"]=strsplit(toString(FS_QA[i,1]),'_')[[1]][1]
  FS_QA[i,"tp"]=strsplit(toString(FS_QA[i,1]),'_')[[1]][2]
}
bmi_group_MRI_QA=merge(bmi_group_MRI,FS_QA, by=c("subj.ID","tp"),all=TRUE)
```

@@

These values describe the average and maximal motion during rsfMRI scans. Used for manuscript on head motion. See further information in `/data/pt_02161/Analysis/Project1_headmotion/`.
```{r "Loading mean/max FD values from /data/pt_02161/Analysis/Preprocessing/qa/rs_qa/ADI_rs_qa.csv", echo=FALSE}
rs_ID = read.csv("/data/pt_02161/Results/Project1_headmotion/subjects_rs.csv")
colnames(rs_ID)[1] = "subj"
res = read.csv(
  "/data/pt_02161/Results/Project1_headmotion/summary_FDvals_based_on_subjects_rs.csv",
  header = FALSE
)
colnames(res) = c("meanFD_bl",
                  "maxFD_bl",
                  "meanFD_fu",
                  "maxFD_fu",
                  "meanFD_fu2",
                  "maxFD_fu2")
rs_FD = data.frame(rs_ID$subj, res)
colnames(rs_FD)[1] = "subj.ID"
rs_FD[rs_FD == 0] = NA

@@

However, because one participant had surgery after the first fu, baseline data will be disregarded and fu and fu2 will be recoded into bl and fu, respectively.

```{r "Reassignment of ADI061 data"}
# remove baseline measure (actually two time points before surgery)
bmi_group_MRI_QA_FD_recoded <- bmi_group_MRI_QA_FD[!(bmi_group_MRI_QA_FD$subj.ID == "ADI061" &
                         bmi_group_MRI_QA_FD$tp == "bl"),]
bmi_group_MRI_QA_FD_recoded$original_tp <- bmi_group_MRI_QA_FD_recoded$tp
# recode fu and fu2 into bl and fu
bmi_group_MRI_QA_FD_recoded$tp[(bmi_group_MRI_QA_FD_recoded$subj.ID == "ADI061" &
                         bmi_group_MRI_QA_FD_recoded$original_tp == "fu")] = "bl"
bmi_group_MRI_QA_FD_recoded$tp[(bmi_group_MRI_QA_FD_recoded$subj.ID == "ADI061" &
                         bmi_group_MRI_QA_FD_recoded$original_tp == "fu2")] = "fu"
```

### Final sample description.
```{r "Some characteristics of this sample", echo=FALSE}
final_sample=bmi_group_MRI_QA_FD_recoded
hist(final_sample$meanFD)
hist(final_sample$Final_Score)
```

>>>>

---
output:
  pdf_document: default
  html_document: default
---

@@

```{r "Package loading", include=FALSE}
library(lme4)
library(ggplot2)
library(plyr)
library(dplyr)
library(tidyr)
library(psych)
library(MuMIn)
```

```{r All info files}
# Messueberblick
Messueberblick_path <- "/data/p_02161/ADI_studie/metadata/ADI_MRT_Messueberblick_2019_04_10.csv"

# ADI_Aktuell  -  main table, BMI info is refilled with Messueberblick entries
ADIAktuell_üath <- "/data/p_02161/ADI_studie/metadata/Daten ADI aktuell.sav"

# QOL_sheet  -  correct group information (unused, just for comparison)
QOLsheet_path <- "/data/pt_02161/Results/Project1_headmotion/QOL_sheet_of_ADI_Messueberblick.csv"

# summary_with_SPSS  - table has errors (contradictions for ADI080 and ADI036 with other table, but contains missing subjects)
GroupCodingWitErrors_path <- "/data/pt_02161/Results/Project1_headmotion/summary_with_SPSS_data_sheet_of_ADI_Messueberblick.csv"

# rs QA infos
rsQA_path <- "/data/pt_02161/Analysis/Preprocessing/qa/rs_qa/ADI_rs_qa.csv"

# FS infos
FSqa_path <- "/data/p_02161/ADI_studie/BIDS/derivatives/freesurfer/QA/freesurfer_qa_qoala_11.7.csv"

# subjects_rs  -  Loading mean/max FD values
rsID_path <- "/data/pt_02161/Results/Project1_headmotion/subjects_rs.csv"
FDvals_path <- "/data/pt_02161/Results/Project1_headmotion/summary_FDvals_correctADI061.csv"

# output
out_path <- "/data/p_02161/ADI_studie/metadata/final_sample_MRI_QA_info.csv"
```
### Loading Demographics
We get BMI from `ADI_MRT_Messueberblick_2019_04.csv` (this is an older version with correct columns (compared to _AW.xlsx). See README in `/data/p_02161/ADI_studie/metadata/`.

@@

```{r "Load data", include=FALSE}
# read "Messüberblick"
# collect subject IDs in one column "BMI"
bmi = read.csv(
  "/data/p_02161/ADI_studie/metadata/ADI_MRT_Messueberblick_2019_04_10.csv",
  na.strings = c("", "NA"),
  stringsAsFactors = FALSE
)[, c("X.1",
      "BMI",
      "X.2",
      "BMI.1",
      "X.3",
      "BMI.2",
      "Geschlecht",
      "Alter")]

bmi$subj.ID = bmi$X.1  
bmi$subj.ID[!is.na(bmi$X.2)] = bmi$X.2[!is.na(bmi$X.2)]  
bmi$subj.ID[!is.na(bmi$X.3)] = bmi$X.3[!is.na(bmi$X.3)]  
bmi = bmi[, c("subj.ID", "BMI", "BMI.1", "BMI.2", "Geschlecht", "Alter")]

# get BMI data from another ADI table which contains more subjects
# select variables: only BMI
bmi_repl = haven::read_sav("/data/p_02161/ADI_studie/metadata/Daten ADI aktuell.sav")[, c(
  "ADI_Code",
  "BMI_BL",
  "BMI_6m",
  "BMI_12m",
  "Sex",
  "Age_BL",
  "Age_6m",
  "Age_12m",
  "Educa_BL"
)]
bmi_repl$subj.ID = gsub("-", "", bmi_repl$ADI_Code)

# replace some missing BMI in the "ADI_aktuell.sav" (bmi_repl) with values from "Messueberblick" (bmi)
merged_bmi = merge(bmi, bmi_repl, by = "subj.ID", all = TRUE)
merged_bmi$BMI_BL[is.na(merged_bmi$BMI_BL)] = merged_bmi$BMI[is.na(merged_bmi$BMI_BL)]
merged_bmi$BMI_6m[is.na(merged_bmi$BMI_6m)] = merged_bmi$BMI.1[is.na(merged_bmi$BMI_6m)]
merged_bmi$BMI_12m[is.na(merged_bmi$BMI_12m)] = merged_bmi$BMI.2[is.na(merged_bmi$BMI_12m)]
merged_bmi$Sex[is.na(merged_bmi$Sex)] = merged_bmi$Geschlecht[is.na(merged_bmi$Sex)]
merged_bmi$Age_BL[is.na(merged_bmi$Age_BL)] = merged_bmi$Alter[is.na(merged_bmi$Age_BL)]
```

```{r "ADI061 correction", include=FALSE}
merged_bmi[merged_bmi$subj.ID == "ADI061",c("BMI_BL","BMI_6m","BMI_12m")]

# indication on what is the correct BMI is in Messueberblick.csv
merged_bmi[merged_bmi$subj.ID == "ADI061",c("BMI_BL","BMI_6m","BMI_12m")] <- 
  c(bmi_repl$BMI_6m[bmi_repl$subj.ID == "ADI061"], # BL = FU value
    bmi$BMI.2[bmi$subj.ID=="ADI061"],              # FU = FU2 value (from ADI_aktuell.sav)
    bmi_repl$BMI_12m[bmi_repl$subj.ID == "ADI061"])# FU2 = FU2 value (from Messueberblick)
merged_bmi[merged_bmi$subj.ID == "ADI061",c("BMI_BL","BMI_6m","BMI_12m")]
```

```{r "long format for data frame", include=FALSE}
# clean up data frame - take refilled BMI of "ADI_aktuell.sav" (bmi_repl)
lbmi <- merged_bmi %>% gather(tp, BMI, BMI_BL, BMI_6m, BMI_12m)
lbmi = lbmi[, c("subj.ID", "Age_BL", "Sex", "tp",  "BMI")]
lbmi$tp = as.factor(lbmi$tp)
levels(lbmi$tp) = c("fu2", "fu", "bl")
lbmi$tp <- factor(lbmi$tp, levels = rev(levels(lbmi$tp)))
lbmi$Sex = as.factor(lbmi$Sex)
levels(lbmi$Sex) = c("male", "female", "male", "female")
```

```{r "Compare merged_bmi with bmi from QOL list", include=FALSE }
# does not use QOL list further on
# compare merged_bmi (ADI_aktuell.sav/ Messueberblick) with bmi from QOL list for those subjects who have it
QOL_sheet = read.csv(
  "/data/pt_02161/Results/Project1_headmotion/QOL_sheet_of_ADI_Messueberblick.csv"
)
QOL_sheet$subj.ID = gsub("-", "", QOL_sheet$X.1)

@@

bmi_compare = merge(QOL_sheet, merged_bmi, by = "subj.ID", all = TRUE)
sum(bmi_compare$BMI.BL - bmi_compare$BMI_BL, na.rm = T)
sum(bmi_compare$BMI.FU - bmi_compare$BMI_6m, na.rm = T)
sum(bmi_compare$BMI.FU_2 - bmi_compare$BMI_12m, na.rm = T) 
# differences because of  ADI061
```

@@

README: "Unfortunately, there are errors regarding BMI assignment (e.g. `/data/p_02161/ADI_studie/metadata/ADI_MRT_Messueberblick_AW.xlsx`, first sheet, where ADI002's FU2 weight is assigned to BL, for example). Similarly, group coding is wrong in the first two sheets (`ADI_MRT_Messueberblick_2019_04_` and `summary with SPSS data`). Correct group information, based on BMI trajectory, seems to be in sheet `QOL` (see ADI080, ADI036).

@@

```{r "Write group coding for all subjects", echo=FALSE}
write.csv(bmi_group[bmi_group$tp=="bl",], "/data/p_02161/ADI_studie/metadata/group_list_all.csv")
```

@@

These infos were gathered during data copying, organizing and preprocessing, especially of the rs data. Yet, infos from other sequences are also included. See `/data/pt_02161/Analysis/Preprocessing/qa/README.md` for further information.

```{r "Loading rs QA infos from /data/pt_02161/Analysis/Preprocessing/qa/rs_qa/ADI_rs_qa.csv", echo=FALSE}
rs_QA=read.csv("/data/pt_02161/Analysis/Preprocessing/qa/rs_qa/ADI_rs_qa.csv")
for (i in 1:nrow(rs_QA)){
  #print(strsplit(toString(rs_QA[i,1]),'_')[[1]][1])
  rs_QA[i,"subj.ID"]=strsplit(toString(rs_QA[i,1]),'_')[[1]][1]
  rs_QA[i,"tp"]=strsplit(toString(rs_QA[i,1]),'_')[[1]][2]
}

@@

See `/data/pt_02161/Analysis/Preprocessing/qa/README.md` for further information.
```{r "Loading FS infos from /data/p_02161/ADI_studie/BIDS/derivatives/freesurfer/QA/freesurfer_qa_qoala_11.7.csv", echo=FALSE}
FS_QA=read.csv("/data/p_02161/ADI_studie/BIDS/derivatives/freesurfer/QA/freesurfer_qa_qoala_11.7.csv")
for (i in 1:nrow(FS_QA)){
  #print(strsplit(toString(rs_QA[i,1]),'_')[[1]][1])
  FS_QA[i,"subj.ID"]=strsplit(toString(FS_QA[i,1]),'_')[[1]][1]
  FS_QA[i,"tp"]=strsplit(toString(FS_QA[i,1]),'_')[[1]][2]
}

```

```{r "Exclude subjects with more than 3 time points"}
# exclude if subject info does not have pattern ADI000_bl ADI000_fu or ADI000_fu2
FS_QA <- FS_QA[stringr::str_detect(FS_QA$Subject, "^ADI\\d{3}_(bl|fu|fu2)$"),]

bmi_group_MRI_QA=merge(bmi_group_MRI,FS_QA, by=c("subj.ID","tp"),all=TRUE)
```

@@

These values describe the average and maximal motion during rsfMRI scans. Used for manuscript on head motion. See further information in `/data/pt_02161/Analysis/Project1_headmotion/`.
```{r "Loading mean/max FD values from /data/pt_02161/Analysis/Preprocessing/qa/rs_qa/ADI_rs_qa.csv", echo=FALSE}
rs_ID = read.csv("/data/pt_02161/Results/Project1_headmotion/subjects_rs.csv")
colnames(rs_ID)[1] = "subj"
res = read.csv(
  "/data/pt_02161/Results/Project1_headmotion/summary_FDvals_correctADI061.csv",
  header = FALSE
)
colnames(res) = c("meanFD_bl",
                  "maxFD_bl",
                  "meanFD_fu",
                  "maxFD_fu",
                  "meanFD_fu2",
                  "maxFD_fu2")
rs_FD = data.frame(rs_ID$subj, res)
colnames(rs_FD)[1] = "subj.ID"
rs_FD[rs_FD == 0] = NA

@@

```{r final sample}
final_sample=bmi_group_MRI_QA_FD
```

### Final sample description.
```{r "Some characteristics of this sample", echo=FALSE}
final_sample=bmi_group_MRI_QA_FD
hist(final_sample$meanFD)
hist(final_sample$Final_Score)
```
